A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-22 inst/manuscript/manuscript.tex
more minor updates

We used a simulation study to investigate performance of four
sampling-analysis combinations: Independent Random Sampling (IRS) with a
design-based analysis, called ``IRS-Design'' (IRS samples are not
spatially balanced because population units are selected independent of
their spatial locations); IRS with a model-based analysis, called
``IRS-Model''; GRTS sampling with a design-based analysis, called
``GRTS-Design''; and GRTS sampling with a model-based analysis, called
``GRTS-Model''. These combinations are also provided in Table
\ref{tab:designanalysis}.

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance is 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Then we recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of times the true mean is contained in its 95\% interval.
The GRTS algorithm and the local neighborhood variance estimator are
available in the \texttt{spsurvey} \textbf{\textsf{R}} package (Dumelle
et al., 2021). FPBK is available in the \texttt{sptotal}
\textbf{\textsf{R}} package (Higham et al., 2021) and covariance
parameters were estimated using Restricted Maximum Likelihood (Harville,
1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. We will use data from the 2012 National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). Specifically, we will analyze mercury concentration in
lakes. Although we can calculate the true mean mercury concentration
values for the 986 lakes from the 2012 NLA, we will explore whether or
not we obtain an adequately precise estimate for the realized mean
mercury concentration if we sample only 100 of the 986 lakes. For each
of the four familiar sampling-analysis combinations (IRS-Design,
IRS-Model, GRTS-Design, and GRTS-Model), we estimate (design-based) or
predict (model-based) the mean mercury concentration and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals
from this sample of 100 lakes, which we compare to the true mean mercury
concentration from all 986 lakes.

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
sampling plan or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
covariance. When there is spatial covariance (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial covariance
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial covariance, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. For all four sampling-analysis combinations, the true
realized mean mercury concentration is within the bounds of the 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Though we should not generalize these results to other samples from
these data, we do note a couple of patterns. The design-based IRS
analysis shows the largest standard error: a likely reason is that this
is the only approach that does not incorporate any spatial information
regarding mercury concentration. Both analyses using GRTS sampling have
lower standard errors than both analyses using IRS sampling. We expect
that these patterns are consistent with other samples from these data
because mercury concentration exhibits spatial dependence, so a
spatially balanced sample should usually yield a lower standard error.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach
incorporates randomness through sampling to estimate population
parameters. The model-based approach incorporates randomness through
distributional assumptions to predict realized values of a stochastic
process. Though these approaches have often been compared in the
literature from theoretical and analytical perspectives, our
contribution lies in studying them in a spatial context while
implementing spatially balanced sampling and the local neighborhood
variance estimator (in the design-based approach). Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, the sampling decision
(IRS vs GRTS) is most important when using a design-based analysis.
Though GRTS-Model still outperformed IRS-Model, the model-based analysis
mitigated much of the inefficiency of the IRS sample. Second,
independent of the analysis approach, we found no reason to prefer IRS
over GRTS for sampling spatial data -- GRTS-Design and GRTS-Model
generally performed at least as well as their IRS counterparts when
there was no spatial covariance and noticeably better than their IRS
counterparts when there was spatial covariance. Third, as the strength
of spatial covariance increases, the gap in rMS(P)E between IRS-Design
and the other sampling-analysis combinations also increases. Fourth and
finally, when the response was normal, interval coverage for all
sampling-analysis combinations was very close to 95\% for all sample
sizes; when the response was lognormal, interval coverage for all
sampling and analysis was between 90\% and 95\% and closest to 95\% when
\(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
also more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and response variable. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the dependence structure in the stochastic process of study. Model
selection is also possible using model-based approaches and criteria
such as cross validation, likelihood ratio tests, or AIC (Akaike, 1974).
Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Model-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations. In short, when deciding
whether the design-based or model-based approach is more appropriate to
implement, the benefits and drawbacks of each approach should be
considered alongside the particular goals of the study.

>>>>

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
is IRS-Design. In IRS-Design, samples are selected using the Independent
Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial
locations of the population units, which implies IRS samples are not
spatially balanced. In IRS-Design, samples are analyzed using the
design-based approach with an IRS variance estimator that does not
incorporate the spatial locations of the units in the sample. The second
sampling-analysis combination is IRS-Model, where samples are selected
using the IRS algorithm and analyzed using the model-based approach via
Restricted Maximum Likelihood (REML) estimation (Harville, 1977;
Patterson and Thompson, 1971; Wolfinger et al., 1994). The third
sampling-analysis combination is GRTS-Design, where samples are selected
using the GRTS algorithm and analyzed using the design-based approach
with the local neighborhood variance estimator. The fourth and final
sampling-analysis combination is GRTS-Model, where samples are selected
using the GRTS algorithm and analyzed using the model-based approach via
REML estimation. These sampling-analysis combinations are also provided
in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS
and GRTS samples, equal inclusion probabilities were assumed for all
population units. When IRS assumes equal inclusion probabilities for all
population units, the algorithm is equivalent to ``simple random
sampling.''

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance was 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Then we recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of times the true mean is contained in its 95\% interval.
The IRS algorithm, IRS variance estimator, GRTS algorithm, and local
neighborhood variance estimator are available in the \texttt{spsurvey}
\textbf{\textsf{R}} package (Dumelle et al., 2021). FPBK is available in
the \texttt{sptotal} \textbf{\textsf{R}} package (Higham et al., 2021).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). We will analyze mercury concentration data collected at
986 lakes as part of the 2012 NLA. Although we can calculate the true
mean mercury concentration values for these 986 lakes, here we will
explore whether or not we can obtain an adequately precise estimate for
the realized mean mercury concentration if we sample only 100 of the 986
lakes. For each of the four familiar sampling-analysis combinations
(IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate
(design-based) or predict (model-based) the mean mercury concentration
and construct 95\% confidence (design-based) or 95\% prediction
(model-based) intervals from this sample of 100 lakes, which we compare
to the true mean mercury concentration from all 986 lakes.

@@

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
algorithm or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
covariance. When there is spatial covariance (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial covariance
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial covariance, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

@@

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. For all four sampling-analysis combinations, the true
realized mean mercury concentration is within the bounds of the 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Though we should not generalize these results to other samples from
these data, we do note a couple of patterns. The design-based IRS
analysis shows the largest standard error: a likely reason is that this
is the only approach that does not incorporate any spatial locations.
Additionally, both analyses using GRTS sampling have lower standard
errors than both analyses using IRS sampling.

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach
incorporates randomness through sampling to estimate population
parameters. The model-based approach incorporates randomness through
distributional assumptions to predict realized values of a stochastic
process. Though these approaches have often been compared in the
literature from theoretical and analytical perspectives, our
contribution lies in studying them in a spatial context while
implementing spatially balanced sampling and the local neighborhood
variance estimator (in the design-based approach). Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, the sampling decision
(IRS vs GRTS) is most important when using a design-based analysis.
Though GRTS-Model still outperformed IRS-Model, the model-based analysis
mitigated most of the inefficiencies that result from the IRS samples
lacking spatial balance. Second, independent of the analysis approach,
we found no reason to prefer IRS over GRTS for sampling spatial data --
GRTS-Design and GRTS-Model generally performed at least as well as their
IRS counterparts when there was no spatial covariance and noticeably
better than their IRS counterparts when there was spatial covariance.
Third, as the strength of spatial covariance increases, the gap in
rMS(P)E between IRS-Design and the other sampling-analysis combinations
also increases. Fourth and finally, when the response was normal,
interval coverage for all sampling-analysis combinations was very close
to 95\% for all sample sizes; when the response was lognormal, interval
coverage for all sampling and analysis was between 90\% and 95\% and
closest to 95\% when \(n = 200\).

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
also more naturally handles binary data, free from the more complicated
logistic regression framework commonly used to analyze binary data in a
model-based approach. The model-based approach, however, can more
naturally quantify the relationship between covariates (predictor
variables) and response variable. The model-based approach also yields
estimated spatial covariance parameters, which help better understand
the dependence structure in the stochastic process of study. Model
selection is also possible using model-based approaches and criteria
such as cross validation, likelihood ratio tests, or AIC (Akaike, 1974).
Model-based approaches are capable of more efficient small-area
estimation than design-based approaches by leveraging distributional
assumptions in areas with few observed sites. Model-based approaches can
also compute site-by-site predictions at unobserved locations and use
them to construct informative visualizations like smoothed maps. In
short, when deciding whether the design-based or model-based approach is
more appropriate to implement, the benefits and drawbacks of each
approach should be considered alongside the particular goals of the
study.
